---
title: "Coursera Regression Models Course Project"
author: "CARLO G. INOVERO"
date: "10/7/2018"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

## Executive Summary
This Coursera Regression Models Course Project aims to know the impact of transmission type on miles per gallon (mpg) for cars in the mtcars dataset. Specifically, it aims to know the following:

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions.

We found out that manual transmission is better than automatic in terms of MPG. However, there is a strong relationship between MPG and other factors like cylinder, weight, displacement and horsepower. 

### Exploratory Analysis

#### The Data Set
The `mtcars` data set  was used for this analysis which was extracted from the 1974 edition of Motor Trend US Magazine and it deals with 1973 - 1974 models. *** Appendix 1 *** describes the variable 
   
```{r results='hide', message=FALSE}
     
      data(mtcars)
      head(mtcars)
    ```

#### Categorical Variables to Factors Convertion
Here, we converted categorical variable to factors
```{r results='hide', message=FALSE}

mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$vs <- as.factor(mtcars$vs)
mtcars$am <- factor(mtcars$am)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
levels(mtcars$am)<-c("Automatic", "Manual")
attach(mtcars)
```

### Statistical Inference

#### T-Test transmission type and MPG
```{r results='hide', message=FALSE}
x_auto_mpg<-mtcars[mtcars$am=="Automatic","mpg"]
y_manual_mpg<-mtcars[mtcars$am=="Manual","mpg"]
```
The p-value is 0.001374 which is below 0.05, thus the T-Test rejects the null hypothesis that the difference between transmission types is 0. Hence, there is a significant difference between automatic and manual transmissions.

```{r results='hide', message=FALSE}
means<-cbind(mean(mtcars[mtcars$am=="Automatic","mpg"]), mean(mtcars[mtcars$am=="Manual","mpg"]))
colnames(means)<-c('Auto','Manual')
rownames(means)<-'Mean'
means
```
The difference estimate between the  transmissions is 7.24494 MPG in favor of manual. T-Test results will be seen at ***Appendix 3 *** Hence, Manual Transmission type is more efficient than automatic.


#### Correlation
Appendix 4a and 4b shows that there are strong relationships between MPG with cylinder, weight, displacement and horsepower.

### REGRESSION MODEL

#### Simple Regression Model
The simple linear model obtained an average of 17.147 MPG for automatic, and an increased of 7.245 MPG for manual transmission. The R square value is 0.3598, thus this model only obtained 36% of the variance.please see *** appendix 5 ***.

```{r results = 'hide', message=FALSE}
    simple_model_mtcars <-lm(mpg~am, data = mtcars)

```

#### Miltivariavate Regression Model

```{r results = 'hide', message=FALSE}

multi_model_mtcars <- lm(mpg~am + cyl + disp + hp + wt, data = mtcars)
```
#### Analysis of Variance using ANOVA
Here, we used anova for simple linear model and multivariate model to confirm that the inclusion of the additional variables provides statistically significant information on the mpg. 

```{r results = 'hide', message=FALSE}
anova(simple_model_mtcars, multi_model_mtcars)
```

The p-value of 8.637e-08 confirms that the inclusion of these variables makes a statistically significant diffference on the outcome of the model. The residuals for non-normality (see *** Appendix 7 ***) are  normally distributed.

### CONCLUSIONS
There is a significant difference in MPG between the manual and automatic transmission type. A manual transmission is more efficient in terms of MPG than automatic car. However, there are other factors affects when determining the efficiency  of MPG of a car. Weight, horsepower, and number of cylinders are more statistically significant to determine the efficiency of MPG of a car.

### APPENDICES

#### Appendix 1 - Variable Description
The data set consists 11 variables 

1. mpg: Miles per US gallon
2. cyl: Number of cylinders
3. disp: Displacement (cubic inches)
4. hp: Gross horsepower
5. drat: Rear axle ratio
6. wt: Weight (lb / 1000)
7. qsec: 1 / 4 mile time
8. vs: V/S
9. am: Transmission (0 = automatic, 1 = manual)
10. gear: Number of forward gears
11. carb: Number of carburetors

#### Appendix 2 - Automatic and Manual Exploratory Box Graph 

The graph showed that there is a significant increase in MPG when for vehicles with a manual transmission vs automatic.

```{r results = 'hide', message=FALSE}
library(ggplot2) #for plots
boxplot(mpg ~ am, data = mtcars, 
        col = (c("purple","green")), 
        main = "MPG by Transmission Type ",
        ylab = "Miles Per Gallon", 
        xlab = "Transmission Type")
```

#### Appendix 3 - T Test for Transmission Type and MPG

```{r results = 'hide', message=FALSE}
t.test(x_auto_mpg, y_manual_mpg)
```


#### Appendix 4a - Correlation Table

```{r results = 'hide', message=FALSE}
library(pander)
data ("mtcars")
mtcars_corr <- cor (mtcars)
panderOptions("digits", 2)
pander(mtcars_corr)
```

#### Appendix 4a - Correlation between MPG and other Variables
```{r results = 'hide', message=FALSE}
pairs(mpg ~ ., data = mtcars)
```

#### Appendix 5 - Simple Regression Model
```{r results = 'hide', message=FALSE}
    summary(simple_model_mtcars)
```

#### Appendix 6 - Multivariate Regression Model
```{r results = 'hide', message=FALSE}
    summary(multi_model_mtcars)
```

#### Appendix 7 - Residual
```{r results = 'hide', message=FALSE}
par(mfrow = c(2, 2))
plot(multi_model_mtcars)
```


